Penalized regression for left-truncated and right-censored survival data

### 2021-01-08 manuscript.tex
update reference style

% bibliography packages
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\bibliographystyle{apa}
\renewcommand{\bibname}{References}

@@

\section{Introduction}
The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), clinical (electronic health records), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information. The amount of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes. Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \citep{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}. Given the hundreds or even thousands of molecular, patho-physiological, and clinical parameters that may be readily available for each patient, developing an appropriate statistical model that is robust to overfitting is critical to ensure good model performance and generalizability. Survival analysis is of particular interest given that overall survival is often a primary outcome of interest.

In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \citep{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model. Similar to ordinary least squares (OLS) regression, these penalized models estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients. This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection. Common penalized approaches include lasso \citep{tibshirani1996regression}, ridge \citep{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \citep{zou2005regularization}, with extensions such as the adaptive lasso \citep{zou2006adaptive}.

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left-truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times. Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \citep{kehl2020selectionbias} and specialized treatments \citep{newman2020immortaltime}. These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.

Left truncation is a well known feature of medical data and there are methods to adjust for it \citep{kalbfleisch2011statistical}. The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \citep{tsai1987note}. Similarly, with left truncation, the likelihood function in a parametric model can be adjusted so that it is conditional on survival to the entry time.

@@

\citet{cox1972regression} proposed a partial likelihood for $\beta$ without involving the baseline hazard $h_0(t)$

@@

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors. In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.

The extension of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \citep{gui2005penalized}, transcriptomes \citep{wu2011penalized}, and injury scale codes \citep{mittal2013penalized}. Like linear regression, the performance and fit of a Cox proportional hazards model is known to deteriorate with large $p$, providing an advantage to penalized approaches in high-dimensional, time-to-event feature sets.

\citet{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for

@@

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by the penalty $\lambda \cdot P_{\alpha}\beta$. As described elsewhere \citep{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \citep{breslow1972} and details provided in \citep{simon2011regularization}.

@@

\subsection{Results}
We used calibration curves \citep{harrell2015regression} to compare models with and without a left truncation adjustment. Each point in the plot consists of simulated patients in a given decile of predicted survival at a given time point. The x-axis is the averaged predicted survival probability and the y-axis is the pseudo observed survival probability computed using the Kaplan-Meier estimator. Perfectly calibrated predictions are those that lie along the 45 degree line so that predicted and observed survival probabilities are equal. The plots consist of patients in the complete sample and are therefore a good test of the effectiveness of the left truncation adjustment.

@@

The results of the large $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities. Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but slightly underestimated for patients with higher survival probabilities. Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.

A common metric used to evaluate survival models is the C-index \citep{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients. We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data. The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data. When using the observed sample to evaluate the model, the C-index is higher in a model that does not adjust for left truncation, even though it is poorly calibrated. Furthermore, even when using the complete sample, the C-indices are very similar.

@@

Both the small and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors. For this reason, model specifications were relatively simple, e.g. modeling linear and additive effects, applying no transformations to predictors (for instance, modeling mutations as binary), and including a variable, year of diagnosis, that is highly correlated with left truncation time but not necessarily with survival (\autoref{fig:left-trunc-time}). Survival was predicted from diagnosis. For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example \citet{lai2020nsclc}.

@@

The larger model included all non-biomarker variables from the smaller model and added nearly all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data to biomarkers with less than 30\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20. Biomarkers with zero variance or that were present in less than 0.1\% of patients were also excluded.

@@

Although not the primary focus of this paper, our approach generated findings that were both consistent with prior literature in genomics and clinically plausible. One of the mutation features that had the highest importance was the presence of short variants in the gene KEAP1. This finding is supported by similar reports of risk-conferring status in an independent lung cancer cohort \citep{shen_harnessing_2019}. Conversely, variants in EGFR and ALK were found to be protective in the CGDB data, putatively because mutations in these genes qualify patients to receive targeted therapies and therefore benefit from improved response and longer survival.

When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \citep{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \citep{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \citep{thiebaut2004choice, chiou2019transformation}.

>>>>

% bibliography packages
\usepackage{natbib}
\setcitestyle{super}
\bibliographystyle{WileyNJD-AMA}
\setcitestyle{open={},close={},comma,sort&compress}
\renewcommand{\bibname}{References}
\renewcommand\bibnumfmt[1]{{#1}.} % to format bibliography as 1. and 2. etc

@@

\section{Introduction}
The recent development and accessibility of genomic (DNA sequencing, microarrays), proteomic (immuno-histology, tissue arrays), clinical (electronic health records), digital (wearables), and imaging (PET, fMRI) technologies have led to an increasing complexity and dimensionality of patient health information. The amount of person-level and disease-specific detail captured by these technologies offers an exciting new opportunity to understand patient history and prognosis by combining these molecular data with macro-scale clinical data on patient characteristics, treatment, and outcomes. Specifically, these data may offer insights into risk factors affecting patient-related outcomes leading to interest in developing prognostic tools to improve clinical decision-making and patient care \cite{gui2005penalized, wishart2010predict, ow2016big, yousefi2017predicting}. Given the hundreds or even thousands of molecular, patho-physiological, and clinical parameters that may be readily available for each patient, developing an appropriate statistical model that is robust to overfitting is critical to ensure good model performance and generalizability. Survival analysis is of particular interest given that overall survival is often a primary outcome of interest.

In a high-dimensional setting, computationally efficient regularization and shrinkage techniques have been developed \cite{tibshirani1997lasso, friedman2010regularization, simon2011regularization} for the Cox model. Similar to ordinary least squares (OLS) regression, these penalized models estimate regression coefficients by minimizing the sum of squared errors, but place a constraint in the equation to minimize the sum of coefficients. This constraint (controlled by a parameter, $\lambda$) penalizes a high number of variables in the model and shrinks the coefficients of less influential features towards zero - acting, in cases when some coefficients are shrunk fully to zero, as automatic feature selection. Common penalized approaches include lasso \cite{tibshirani1996regression}, ridge \cite{hoerl1970ridgeA, hoerl1970ridgeB}, and elastic net \cite{zou2005regularization}, with extensions such as the adaptive lasso \cite{zou2006adaptive}.

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left-truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \cite{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times. Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \cite{kehl2020selectionbias} and specialized treatments \cite{newman2020immortaltime}. These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.

Left truncation is a well known feature of medical data and there are methods to adjust for it \cite{kalbfleisch2011statistical}. The Kaplan-Meier estimator of the survival function requires almost no modifications as it is only necessary to change the risk set so that individuals are only at risk of an event after study entry \cite{tsai1987note}. Similarly, with left truncation, the likelihood function in a parametric model can be adjusted so that it is conditional on survival to the entry time.

@@

Cox\cite{cox1972regression} proposed a partial likelihood for $\beta$ without involving the baseline hazard $h_0(t)$

@@

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \cite{tibshirani1996regression} and $l_2$ (ridge) \cite{tikhonov1963ridge} regression, elastic net regression \cite{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \cite{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors. In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.

The extension of penalized regression methods to survival data have been extensively described and applied, for example in microarray gene expression data \cite{gui2005penalized}, transcriptomes \cite{wu2011penalized}, and injury scale codes \cite{mittal2013penalized}. Like linear regression, the performance and fit of a Cox proportional hazards model is known to deteriorate with large $p$, providing an advantage to penalized approaches in high-dimensional, time-to-event feature sets.

Simon et. al\cite{simon2011regularization} introduced a fast, pathwise algorithm to fit a regularized Cox proportional hazards model solving for

@@

where $\beta$ is the vector of regression coefficients and is maximized over \autoref{eqn:simon} subject to a constraint defined by the penalty $\lambda \cdot P_{\alpha}\beta$. As described elsewhere \cite{simon2011regularization}, the penalty can take the form of the $l_1$ (lasso), $l_2$ (ridge), or elastic net penalty,  and the algorithm fits over a path of $\lambda$ values via cyclical coordinate descent. Instead of solving a least squares problem at each iteration, the algorithm solves a penalized reweighted least squares problem, in which observation weights $w$ are defined in the coordinate descent least squares minimization step. The optimal $\lambda$ can be identified through cross-validation. Tied survival times are handled via the Breslow approximation \cite{breslow1972} and details provided in \cite{simon2011regularization}.

@@

\subsection{Results}
We used calibration curves \cite{harrell2015regression} to compare models with and without a left truncation adjustment. Each point in the plot consists of simulated patients in a given decile of predicted survival at a given time point. The x-axis is the averaged predicted survival probability and the y-axis is the pseudo observed survival probability computed using the Kaplan-Meier estimator. Perfectly calibrated predictions are those that lie along the 45 degree line so that predicted and observed survival probabilities are equal. The plots consist of patients in the complete sample and are therefore a good test of the effectiveness of the left truncation adjustment.

@@

The results of the large $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities. Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but slightly underestimated for patients with higher survival probabilities. Probability rescaling may be effective in such cases, for example with the use of Platt scaling \cite{platt1999scaling} or isotonic regression \cite{niculescu2005isotonic}.

A common metric used to evaluate survival models is the C-index \cite{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients. We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data. The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data. When using the observed sample to evaluate the model, the C-index is higher in a model that does not adjust for left truncation, even though it is poorly calibrated. Furthermore, even when using the complete sample, the C-indices are very similar.

@@

Both the small and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors. For this reason, model specifications were relatively simple, e.g. modeling linear and additive effects, applying no transformations to predictors (for instance, modeling mutations as binary), and including a variable, year of diagnosis, that is highly correlated with left truncation time but not necessarily with survival (\autoref{fig:left-trunc-time}). Survival was predicted from diagnosis. For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example Lai and colleagues\cite{lai2020nsclc}.

@@

The larger model included all non-biomarker variables from the smaller model and added nearly all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data to biomarkers with less than 30\% missingness and imputed the remaining mutation data using k nearest neighbors \cite{troyanskaya_missing_2001}, with k = 20. Biomarkers with zero variance or that were present in less than 0.1\% of patients were also excluded.

@@

Although not the primary focus of this paper, our approach generated findings that were both consistent with prior literature in genomics and clinically plausible. One of the mutation features that had the highest importance was the presence of short variants in the gene KEAP1. This finding is supported by similar reports of risk-conferring status in an independent lung cancer cohort \cite{shen_harnessing_2019}. Conversely, variants in EGFR and ALK were found to be protective in the CGDB data, putatively because mutations in these genes qualify patients to receive targeted therapies and therefore benefit from improved response and longer survival.

When making predictions with survival data is of interest, it is critical to define the risk set appropriately, as our work demonstrates. To do this requires recognizing the underlying process by which individuals are captured in the data in the first place; i.e., when and how individuals came to be included in the dataset of interest. One of the motivations for the current work derives from the advent of biobank and genomic test databases that store information on patient illness - presumably after the patient has been diagnosed and ``at risk'' for some time. In these data, overlooking the ``delayed entry'' of at-risk patients could result in misleading conclusions, as we demonstrate in this work. Additionally, patient cohorts marked by a milestone entry, such as genomic testing, may not only be subject to immortal time bias but also to a temporal selection bias. The latter may occur if left truncation time is correlated with survival time, for example, if genomic testing is ordered selectively for patients with worse prognosis. This was uncovered in a cohort of multi-stage cancer patients who received genomic testing \cite{kehl2020selectionbias} and can affect inference even in the presence of a left truncation adjustment. While the cohort of lung cancer patients in our real-world use case was restricted to stage IV diagnoses, mitigating this potential selection bias \cite{kehl2020selectionbias}, it can be adjusted for by modeling the association between left truncation time and survival \cite{thiebaut2004choice, chiou2019transformation}.
